Collective SU($n$) spin model dynamics

### 2020-08-18 sun_phases/multilevel_spin_notes.tex
add some drive hamiltonians
add more drive Hamiltonians
mionr fixes

Rather than coupling the states of a multilevel system directly with a
single driving laser, as in Section \ref{sec:drive_SOC}, we now
consider coupling through an off-resonant Raman transition to a highly
excited state.  Specifically, we consider a three-laser drive of the
form

@@

where $j$ indexes a lattice site, $\Omega_m$ is a (real) driving
amplitude; $\phi_m$ is the relative phase of drive $m$ between
adjacent lattice sites; $\op{r}{s}$ for $r,s\in\set{\up,\dn}$ is a
pseudo-spin operator for the auxiliary degree of freedom that is
off-resonantly addressed by the lasers; $\Delta$ is a detuning of the
lasers from the energy of an excited state; and $\1$ is the identity
operator.  If the detuning is large, $\abs{\Delta}\gg\abs{\Omega_m}$,
then at second order in perturbation theory the effective action of
$H_{\t{Raman,bare}}^{(\phi)}$ within the $\ket{\dn}$-state manifold
takes the form

@@

where $H_{\t{Raman},\tau}^{(\phi)}$ generates nuclear spin transitions
$\mu\to\mu\pm\tau$:

@@

where the coefficients $\Omega_{LM}^{(\phi,j)}$ are provided in
Appendix \ref{sec:drive_raman_coeff}.  If the driving lasers are
phase-matched with $\phi_m=m\phi$, such that a nuclear spin transition
$\mu\to\mu+m$ on site $j$ imprints the phase $m\phi j$, then the drive
$H_{\t{Raman}}^{(\phi)}$ takes the form

@@

with rotated collective drive operators $\D_{LM}^{(\phi)}$ defined in
\eqref{eq:drive_rot} and uniform driving amplitudes $\Omega_{LM}$
provided in Appendix \ref{sec:drive_raman_coeff}.  Similarly to the
case of a direct drive considered in Section \ref{sec:drive_SOC}, the
Raman drive $H_{\t{Raman}}^{(\phi)}$ in \eqref{eq:drive_raman_matched}
can be made homogenous through the gauge transformation
$c_{j\mu}^\dag \to e^{-i\mu\phi j} c_{j\mu}^\dag$, which takes

@@

Writing $H_{\t{Raman}} = \sum_j H_{\t{Raman}}^{\t{single},j}$, where
$H_{\t{Raman}}^{\t{single},j}$ denotes the action of
$H_{\t{Raman}}^{\t{single}}$ on spin $j$, we can expand

@@

where
$\p{S_\z,S_\x,S_\y} \equiv \xi_1\times\p{D_{1,0},D_{1,1},D_{1,-1}}$ is
the standard SU(2) subalgebra of the operators on a single $n$-level
spin, and

@@

Tuning the amplitudes $\Omega_0,\Omega_+,\Omega_-$, or equivalently
$h_0,h_+,h_-$, enables us to implement various drive Hamiltonians,
several of which are shown in Table \ref{tab:drives}.

\begin{table}
  \centering
  \caption{Drive Hamiltonians that can be implemented with various
    amplitude-matching conditions.}
  \begin{tabular}{c|c|c}
    Condition & $\p{h_0,h_+,h_-}/h$
    & $H_{\t{Raman}}^{\t{single}}/\p{h^2/\Delta}$ \\ \hline\hline
    $\Omega_0=\Omega_+=+\Omega_-$ & $\p{1,\sqrt{2},0}$
    & $\sqrt{2} S_\x + S_\x^2 - S_\y^2$ \\ \hline
    $\Omega_0=\Omega_+=-\Omega_-$ & $\p{1,0,\sqrt{2}}$
    & $\sqrt{2}\p{S_\z S_\x + S_\x S_\z} - \p{S_\x^2-S_\y^2}$ \\ \hline
    $\Omega_+=\Omega_-=0$ & $\p{1,0,0}$
    & $S_\x^2 + S_\y^2$ \\ \hline
    $\Omega_0=\Omega_-=0$ & $\p{0,1,+1}$
    & $+S_\z - \p{S_\x^2 + S_\y^2}$ \\ \hline
    $\Omega_0=\Omega_+=0$ & $\p{0,1,-1}$
    & $-S_\z - \p{S_\x^2 + S_\y^2}$
  \end{tabular}
  \label{tab:drives}
\end{table}

>>>>

Rather than coupling the states of a multilevel system directly with a single driving laser, as in Section \ref{sec:drive_SOC}, we now consider coupling through an off-resonant Raman transition to a highly excited state.
Specifically, we consider a three-laser drive of the form

@@

where $j$ indexes a lattice site, $\Omega_m$ is a (real) driving amplitude; $\phi_m$ is the relative phase of drive $m$ between adjacent lattice sites; $\op{r}{s}$ for $r,s\in\set{\up,\dn}$ is a pseudo-spin operator for the auxiliary degree of freedom that is off-resonantly addressed by the lasers; $\Delta$ is a detuning of the lasers from the energy of an excited state; and $\1$ is the identity operator.
If the detuning is large, $\abs{\Delta}\gg\abs{\Omega_m}$, then at second order in perturbation theory the effective action of $H_{\t{Raman,bare}}^{(\phi)}$ within the $\ket{\dn}$-state manifold takes the form

@@

where $H_{\t{Raman},\tau}^{(\phi)}$ generates nuclear spin transitions $\mu\to\mu\pm\tau$:

@@

where the coefficients $\Omega_{LM}^{(\phi,j)}$ are provided in Appendix \ref{sec:drive_raman_coeff}.
If the driving lasers are phase-matched with $\phi_m=m\phi$, such that a nuclear spin transition $\mu\to\mu+m$ on site $j$ imprints the phase $m\phi j$, then the drive $H_{\t{Raman}}^{(\phi)}$ takes the form

@@

with rotated collective drive operators $\D_{LM}^{(\phi)}$ defined in \eqref{eq:drive_rot} and uniform driving amplitudes $\Omega_{LM}$ provided in Appendix \ref{sec:drive_raman_coeff}.
Similarly to the case of a direct drive considered in Section \ref{sec:drive_SOC}, the Raman drive $H_{\t{Raman}}^{(\phi)}$ in \eqref{eq:drive_raman_matched} can be made homogenous through the gauge transformation $c_{j\mu}^\dag \to e^{-i\mu\phi j} c_{j\mu}^\dag$, which takes

@@

Writing $H_{\t{Raman}} = \sum_j H_{\t{Raman}}^{\t{single},j}$, where $H_{\t{Raman}}^{\t{single},j}$ denotes the action of $H_{\t{Raman}}^{\t{single}}$ on spin $j$, we can expand

@@

where $\p{S_\z,S_\x,S_\y} \equiv \xi_1\times\p{D_{1,0},D_{1,1},D_{1,-1}}$ is the standard $\mathfrak{su}(2)$ subalgebra of the operators on a single $n$-level spin, and

@@

Tuning the amplitudes $\Omega_0,\Omega_+,\Omega_-$, or equivalently $h_0,h_+,h_-$, enables us to implement various drive Hamiltonians, several of which are shown in Table \ref{tab:drives}.

\begin{table}
  \centering
  \caption{Drive Hamiltonians that can be implemented with various amplitude-matching conditions.}
  \begin{tabular}{c|c|c}
    Condition & $\p{h_0,h_+,h_-}/h$
    & $H_{\t{Raman}}^{\t{single}}/\p{h^2/\Delta}$
    \\ \hline\hline
    $\Omega_0=\pm\Omega_+=\pm\Omega_-$ & $\p{1,\pm\sqrt{2},0}$
    & $\pm\sqrt{2} S_\x + \p{S_\x^2 - S_\y^2}$
    \\ \hline
    $\Omega_0=\pm\Omega_+=\mp\Omega_-$ & $\p{1,0,\pm\sqrt{2}}$
    & $\pm\sqrt{2}\p{S_\z S_\x + S_\x S_\z} - \p{S_\x^2-S_\y^2}$
    \\ \hline
    $\Omega_0=\Omega_\mp=0$ & $\p{0,1,\pm1}$
    & $\pm S_\z + S_\z^2$
    \\ \hline
    $\sqrt{2}\Omega_0=\pm\Omega_+=\pm\Omega_-$
    & $\p{1,\pm1,0}$ & $\pm S_\x + S_\x^2$
    \\ \hline
    $\sqrt{2}\Omega_0=\pm\Omega_+=\mp\Omega_-$
    & $\p{1,0,\pm1}$ & $\pm\p{S_\z S_\x+S_\x S_\z} + S_\y^2$
    \\ \hline
    $\Omega_+=\Omega_-=0$ & $\p{1,0,0}$ & $-S_\z^2$
    \\ \hline
    $\Omega_0=0$,  $\Omega_+=+\Omega_-$ & $\p{0,1,0}$ & $-S_\y^2$
    \\ \hline
    $\Omega_0=0$,  $\Omega_+=-\Omega_-$ & $\p{0,0,1}$ & $-S_\x^2$
  \end{tabular}
  \label{tab:drives}
\end{table}

### 2020-08-18 sun_phases/sun_dynamics_notes.tex
mionr fixes

We first consider a general quadratic spin Hamiltonian of the form
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h^{\mu\nu i}_{\rho\sigma j} s_{\mu\nu i} s_{\rho\sigma j}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  \label{eq:spin}
\end{align}
where $\mu,\nu\in\set{I,I-1,\cdots,-I}$ with $I\equiv\p{n-1}/2$ index orthogonal states of an $n$-level spin; $i,j\in\ZZ_N$ index one of $N$ spins; $h^{\mu\nu i}_{\rho\sigma j}$ and $\epsilon_{\mu\nu i}$ are scalars; and $s_{\mu\nu i}\coloneqq\op{\mu}{\nu}_i$ is a transition operator for spin $i$.
If interactions are SU($n$) symmetric then $h^{\mu\nu i}_{\rho\sigma k} = h_{ik} \delta_{\mu\sigma}\delta_{\nu\rho}$, in which case
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h_{ij} \Pi_{ij}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  &&
  \Pi_{ij} \equiv \sum_{\mu,\nu} s_{\mu\nu i} s_{\rho\sigma j}.
  \label{eq:spin_sun}
\end{align}

@@

We consider the single-particle Hamiltonian
\begin{align}
  H = -\f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\varphi} s_{\mu\mu q},
\end{align}
where $q$ is a quasi-momentum (in units with lattice spacing $a=1$); $\mu\in\set{I,I-1,\cdots,-I}$ is a nuclear spin projection onto a quantization axis, with $I\equiv\p{n-1}/2$ the total nuclear spin; $\varphi$ is a spin-orbit-coupling angle, and $J$ is the single-particle bandwidth (proportional to the tunneling rate, $J/4$).
As dynamical order parameters, we consider the operators
\begin{align}
  s_{\mu\nu} \equiv \f1N \sum_q s_{\mu\nu q},
  &&
  s^2 \equiv \v s^\dag\cdot \v s
  = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q},
\end{align}
where $\v s_q\equiv\sum_{\mu,\nu} s_{\mu\nu q}\uv e_{\mu\nu}$ is a vector of normalized generators of $U(n)$ for spin $q$, and $\uv e_{\mu\nu}$ is a unit vector of $\RR^{n^2}$.
In the Heisenberg picture, these operators at time $t$ are
\begin{align}
  s_{\mu\nu}\p{t} = \f1N \sum_q s_{\mu\nu q}
  \times e^{\i t J \eta_{\mu\nu q}},
  &&
  s^2\p{t} = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q}
  \times e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}},
\end{align}
where
\begin{align}
  \eta_{\mu\nu q}
  \equiv -\f12 \sp{\cos\p{q+\mu\varphi} - \cos\p{q+\nu\varphi}}
  = \sin\p{\varphi_{\mu\nu}^-} \sin\p{p+\varphi_{\mu\nu}^+},
  &&
  \varphi_{\mu\nu}^\pm \equiv \f{\mu\pm\nu}{2}\,\varphi.
\end{align}
In particular, we are interested in the expectation values
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  = \f1N \bk{\psi\op{\mu}{\nu}\psi} \sum_q e^{\i t J \eta_{\mu\nu q}}
  = \bk{\psi\op{\mu}{\nu}\psi} \chi_{\mu\nu}\p{\varphi},
\end{align}
and
\begin{align}
  \bk{\Psi|s^2\p{t}|\Psi}
  &= \f1{N^2} \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p\ne q} e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \f{n}{N} \\
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \abs{\chi_{\mu\nu}\p{\varphi}}^2 + \f{n-1}{N},
\end{align}
where
\begin{align}
  \chi_{\mu\nu}\p{\varphi} \equiv \f1N \sum_q e^{\i tJ\eta_{\mu\nu q}}
  = \f1N \sum_q e^{\i tJ\sin\p{\varphi_{\mu\nu}^-} \sin\p{q+\varphi_{\mu\nu}^+}}.
\end{align}
As $N\to\infty$, this sum is well approximated by an integral:
\begin{align}
  \chi_{\mu\nu}\p{\varphi} \approx \f1{2\pi} \int_{-\pi}^\pi \dd\theta\,
  e^{\i tJ\sin\p{\varphi_{\mu\nu}^-} \sin\p{\theta+\varphi_{\mu\nu}^+}}
  = \J_0\sp{tJ\sin\p{\varphi_{\mu\nu}^-}},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
To leading order in $N\gg1$, we thus find that
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  &\approx \bk{\psi\op{\mu}{\nu}\psi}
  \J_0\sp{tJ\sin\p{\varphi_{\mu\nu}^-}},
  \\
  \bk{\Psi|s^2\p{t}|\Psi}
  &\approx \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \J_0\sp{tJ\sin\p{\varphi_{\mu\nu}^-}}^2.
  \label{eq:ops_time}
\end{align}

@@

An empirical estimate $\tilde\rho$ of $\rho$ is then\footnote{This estimate can be improved by defining $\tilde\rho$ as the least-squares fit to the set of linear equations $\tr\p{\tilde\rho\,\Pi_j} = M_j$, and performing maximum-likelihood corrections to $\tilde\rho$ \cite{smolin2012efficient}, which essentially consists of
\begin{enumerate*}
\item diagonalizing $\tilde\rho$,
\item eliminating the most negative eigenvalue (setting it to zero),
\item adding an equal amount to all other eigenvalues to enforce $\tr\tilde\rho=1$, and
\item repeating steps (ii,iii) until there are no more negative eigenvalues.
\end{enumerate*}
This algorithm finds the closest density operator to $\tilde\rho$ that has no negative eigenvalues, where distance is measured by the metric induced by the operator 2-norm: $\norm{\O}_2\equiv\sqrt{\tr\p{\O^\dag \O}}$.}

@@

Indexing normalized eigenvectors and the corresponding eigenvalues of $B_V$ by $\alpha$, respectively $\v x^V_\alpha$ and $m^V_\alpha$, for all $\alpha$ with eigenvalue $m^V_\alpha\ne0$ we can then construct the normalized and linearly independent operators

@@

The amount of information that $M_V$ contains about $\bk{Q^V_\alpha}_\rho$ is controlled, in part, by the magnitude of the eigenvalue $m^V_\alpha$: if $m^V_\alpha$ is large, then $M_V$ will generally contain a lot of information about $\bk{Q^V_\alpha}_\rho$, and vice versa\footnote{Think about the limit $m^V_\alpha\to0$, in which case the fact that $Q^V_\alpha$ is normalized implies that the operator $\sum_j x^V_{\alpha j} \Pi^V_j$ must have a vanishing norm.
  This vanishing norm can only occur due to cancellations that are analogous to a ``sign problem,'' making the expectation value of $\sum_j x^V_{\alpha j} \Pi^V_j$ difficult to estimate because it will be dominated by measurement noise.}.

>>>>

We first consider a general quadratic spin Hamiltonian of the form
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h^{\mu\nu i}_{\rho\sigma j} s_{\mu\nu i} s_{\rho\sigma j}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  \label{eq:spin}
\end{align}
where $\mu,\nu\in\set{I,I-1,\cdots,-I}$ with $I\equiv\p{n-1}/2$ index orthogonal states of an $n$-level spin; $i,j\in\ZZ_N$ index one of $N$ spins; $h^{\mu\nu i}_{\rho\sigma j}$ and $\epsilon_{\mu\nu i}$ are scalars; and $s_{\mu\nu i}\coloneqq\op{\mu}{\nu}_i$ is a transition operator for spin $i$.
If interactions are SU($n$) symmetric then $h^{\mu\nu i}_{\rho\sigma k} = h_{ik} \delta_{\mu\sigma}\delta_{\nu\rho}$, in which case
\begin{align}
  H = \f1N \sum_{i<j}
  h_{ij} \Pi_{ij}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  &&
  \Pi_{ij} \equiv \sum_{\mu,\nu} s_{\mu\nu i} s_{\nu\mu j}.
  \label{eq:spin_sun}
\end{align}

@@

We consider the single-particle Hamiltonian
\begin{align}
  H = -\f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\varphi} s_{\mu\mu q},
\end{align}
where $q$ is a quasi-momentum (in units with lattice spacing $a=1$); $\mu\in\set{I,I-1,\cdots,-I}$ is a nuclear spin projection onto a quantization axis, with $I\equiv\p{n-1}/2$ the total nuclear spin; $\varphi$ is a spin-orbit-coupling angle, and $J$ is the single-particle bandwidth (proportional to the tunneling rate, $J/4$).
As dynamical order parameters, we consider the operators
\begin{align}
  s_{\mu\nu} \equiv \f1N \sum_q s_{\mu\nu q},
  &&
  s^2 \equiv \v s^\dag\cdot \v s
  = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q}.
\end{align}
In the Heisenberg picture, these operators at time $t$ are
\begin{align}
  s_{\mu\nu}\p{t} = \f1N \sum_q s_{\mu\nu q}
  \times e^{\i t J \eta_{\mu\nu q}},
  &&
  s^2\p{t} = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q}
  \times e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}},
\end{align}
where
\begin{align}
  \eta_{\mu\nu q}
  \equiv -\f12 \sp{\cos\p{q+\mu\varphi} - \cos\p{q+\nu\varphi}}
  = \sin\p{\varphi_{\mu\nu}^-} \sin\p{q+\varphi_{\mu\nu}^+},
  &&
  \varphi_{\mu\nu}^\pm \equiv \f{\mu\pm\nu}{2}\,\varphi.
\end{align}
In particular, we are interested in the expectation values
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  = \f1N \bk{\psi\op{\mu}{\nu}\psi} \sum_q e^{\i t J \eta_{\mu\nu q}}
  = \bk{\psi\op{\mu}{\nu}\psi} \chi_{\mu\nu}\p{\varphi},
\end{align}
and
\begin{align}
  \bk{\Psi|s^2\p{t}|\Psi}
  &= \f1{N^2} \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p\ne q} e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \f{n}{N} \\
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \abs{\chi_{\mu\nu}\p{\varphi}}^2 + \f{n-1}{N},
\end{align}
where
\begin{align}
  \chi_{\mu\nu}\p{\varphi} \equiv \f1N \sum_q e^{\i tJ\eta_{\mu\nu q}}
  = \f1N \sum_q e^{\i tJ\sin\p{\varphi_{\mu\nu}^-} \sin\p{q+\varphi_{\mu\nu}^+}}.
\end{align}
As $N\to\infty$, this sum is well approximated by an integral:
\begin{align}
  \chi_{\mu\nu}\p{\varphi} \approx \f1{2\pi} \int_{-\pi}^\pi \dd\theta\,
  e^{\i tJ\sin\p{\varphi_{\mu\nu}^-} \sin\p{\theta+\varphi_{\mu\nu}^+}}
  = \J_0\sp{tJ\sin\p{\varphi_{\mu\nu}^-}},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
To leading order in $N\gg1$, we thus find that
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  &\approx \bk{\psi\op{\mu}{\nu}\psi}
  \J_0\sp{tJ\sin\p{\varphi_{\mu\nu}^-}},
  \\
  \bk{\Psi|s^2\p{t}|\Psi}
  &\approx \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \J_0\sp{tJ\sin\p{\varphi_{\mu\nu}^-}}^2.
  \label{eq:ops_time}
\end{align}

@@

An empirical estimate $\tilde\rho$ of $\rho$ is then\footnote{This estimate can be improved by defining $\tilde\rho$ as the least-squares fit to the set of linear equations $\tr\p{\tilde\rho\,\Pi_j} = M_j$, and performing maximum-likelihood corrections to $\tilde\rho$ \cite{smolin2012efficient}.}

@@

Indexing the (normalized) eigenvectors and the corresponding eigenvalues of $B_V$ by $\alpha$, respectively $\v x^V_\alpha$ and $m^V_\alpha$, for all $\alpha$ with eigenvalue $m^V_\alpha\ne0$ we can then construct the normalized and linearly independent operators

@@

The amount of information that $M_V$ contains about $\bk{Q^V_\alpha}_\rho$ is determined, in part, by the magnitude of the eigenvalue $m^V_\alpha$: if $m^V_\alpha$ is small, then $M_V$ will generally contain little information about $\bk{Q^V_\alpha}_\rho$, and vice versa\footnote{Think about the limit $m^V_\alpha\to0$, in which case the fact that $Q^V_\alpha$ is normalized implies that the operator $\sum_j x^V_{\alpha j} \Pi^V_j$ must have a vanishing norm.
  A vanishing norm can only occur due to fine-tuned cancellations that make the expectation value of $\sum_j x^V_{\alpha j} \Pi^V_j$ difficult to estimate due to the presence of measurement errors and shot noise.}.
